---
title: "A Case Study from 2022 IPUMS USA"
subtitle: "Analyzing state-level populations using the ratio estimation method reveals critical differences in population distribution."
author: 
  - Yuxuan Wei
thanks: "Code and data are available at: https://github.com/wyx827/IPUMS_EDUC"
date: today
date-format: long
format: pdf
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

# Load libraries
library(dplyr)
library(tidyr)
library(tidyverse)

# Read the data
data <- read.csv("/Users/shaw/Desktop/STA304/Exercise/usa_00002.csv")

# Convert relevant columns to factors
data <- data %>%
  mutate(across(c(STATEICP, EDUCD), as.factor))
```

# Introduction

This study utilizes the 2022 American Community Survey (ACS) dataset, obtained from the Integrated Public Use Microdata Series USA [@ipumsusa]. The primary goal is to analyze the relationship between educational attainment, specifically doctoral degrees, and state-level population estimates using the **ratio estimation method**.

The ratio estimation method provides a way to generalize known relationships within a sample (e.g., California) to estimate populations across states. By understanding the proportion of doctoral degree holders in California and applying this to other states, we estimate total populations. This approach reveals both the utility and limitations of ratio estimators in population studies.

Data are extract from IPUMS_USA. Two variables were extracted from the 2022 ACS: STATEICP and EDUC. After selecting the correct data and converting the file from a .dat to a .dta, you could download the data extract from IPUMS USA. You then unzip the file and import it into R [@citeR].This analysis uses the dplyr, tidyr, and haven packages in order to synthesize the data [@tidyverse].

# Doctoral Degree Holders by State
In @tbl-doctoral, we observe the number of doctoral degree holders by state. For example, state code 3 (California) has the highest number of doctoral degree holders with 2014, while states like 2 (Massachusetts) and 6 (Indiana) have significantly fewer, with 165 and 131, respectively. These differences highlight the uneven distribution of doctoral degree holders across states.

The variation in these numbers reflects the unique characteristics of each state. For example

- States like California (STATEICP = 3) and Florida (STATEICP = 12) have large populations, leading to a higher count of doctoral degree holders.

- In contrast, smaller states like Indiana (STATEICP = 6) show lower numbers due to their smaller populations or lower rates of educational attainment.

This foundational data allows us to compute ratios, which we later use for estimating the total population in each state.

```{r}
#| label: tbl-doctoral
#| tbl-cap: "Doctoral Degree Holders by State"
#| echo: false
#| warning: false
#| message: false

# Filter out respondents with a doctoral degree
doctoral_respondents <- data %>% filter(EDUCD == 116)

# Summarize the number of doctoral degree holders by state
doctoral_count_by_state <- doctoral_respondents %>%
  group_by(STATEICP) %>%
  summarise(doctoral_count = n(), .groups = 'drop')

# Display the first 8 rows of the summarized data in table format
knitr::kable(head(doctoral_count_by_state, 8), format = "simple", col.names = c('State Code', 'Doctoral Degree Holders'))
```

# Ratio Estimation for State Populations
The ratio estimation method calculates a baseline proportion by using California's population data, where both the total number of respondents and the number of doctoral degree holders are known. The ratio is then applied to other states to estimate their total populations based on the number of doctoral degree holders in each state.

In @tbl-population-estimate, we compare the estimated total population with the actual total population for a few states:

- California (State Code = 3), which serves as our baseline, has an estimated total of 124,340 and an actual total of 73,077. This overestimation is likely because the proportion of doctoral degree holders in California is used as a reference, and California may have different characteristics compared to other states.

- Massachusetts (State Code = 2) has an estimated total of 10,187 compared to an actual total of 14,523. This discrepancy indicates that Massachusetts has a different ratio of doctoral degree holders relative to its population than California, leading to an underestimation.

- For New York (State Code = 13), the estimated total (174,656) is lower than the actual total (203,891). This suggests that New York's proportion of doctoral degree holders differs significantly from California's, likely due to regional differences in education levels and demographics.

These examples illustrate that the ratio estimation method is sensitive to the underlying assumptions about the relationship between doctoral degree holders and the general population across states. Significant variations in educational attainment levels between states can lead to discrepancies between estimated and actual populations, emphasizing the need for careful consideration when applying ratio estimations to diverse regions.

```{r}
#| label: tbl-population-estimate
#| tbl-cap: "Estimated vs Actual Population Using Ratio Estimation"
#| echo: false
#| warning: false
#| message: false

# Define California's total population and doctoral degree holders
california_total <- 391171
california_doctoral <- doctoral_count_by_state %>%
  filter(STATEICP == "71") %>%
  pull(doctoral_count) # Pull 'doctoral_count' for California

# Calculate the ratio
california_ratio <- california_doctoral / california_total

# Actual total respondents in each state
actual_total_by_state <- data %>%
  group_by(STATEICP) %>%
  summarise(actual_total = n(), .groups = 'drop')

# Estimate total population and combine with actual data
comparison <- doctoral_count_by_state %>%
  transmute(STATEICP, estimated_total = round(doctoral_count / california_ratio)) %>%
  left_join(actual_total_by_state, by = "STATEICP")

# Display the first few rows
knitr::kable(head(comparison, 10), format = "simple", col.names = c('State Code', 'Estimated Total', 'Actual Total'))

```

\newpage

## References {#bibliography}
